---
title: "Analyses of archive & monitoring data "
subtitle: "MUC GoF & Gotland & Landsort"
author: "Juliane Romahn"
date: "07.12.23"
output:
  pdf_document: default
  html_document: default
  word_document: default
editor_options:
  chunk_output_type: inline
papersize: a4
---

```{r}
rm(list = ls())
library(tidyverse)
library(ggplot2)
library(ggpubr)
#library(viridis)
library(nord)
# install.packages("devtools")
#devtools::install_github("G-Thomson/Manu")
library(Manu)
#library(gapminder) 
```

#load input data and prepare output


```{r setup}
# has to be in setup & is only correct outside of the setup
knitr::opts_knit$set(root.dir = '/Users/juliane/Documents/00_Work_SGN/00_PhytoArk/XX_PAPERS/2025_Comparison_Paper__relAbudance/')
getwd()
```

```{r}
source("02_Scripts/00_R_functions.R") # read file with functions

# input path of community & biomonitoring data
input_path <-"02_intermediate"
metadata_path <- "03_metadata_final"

#load biomonitoring data
load(file=file.path(input_path,"ICES_phytodata_with_worms_EMB262MUCs__SprSum.RData"))
ices_data <- muc_ICES_data
rm(muc_ICES_data)

#load community and taxonomy
#load( file=file.path(input_path_seda,"sedaDNA_taxonomy_eukaryotes_cleaned.RData"))
load(file=file.path(input_path,"sedaDNA_community_eukaryotes_normed500_transformed.RData"))
seda_community <- seda_community_sum%>%ungroup() %>% mutate(sample_id=NULL)
rm(seda_community_sum)
unique(gsub("C.*", "C",seda_community$tag))

#load toc (überarbeiten, veraltete datei)
meta_file <- file.path(input_path,"EMB262_PhytoArk_MUC__detailed_metadata_samples.tsv")
TOC_data <- read.table(meta_file, sep="\t", header=T)
TOC_data <- TOC_data[,c("station", "depth", "age" ,"TOC")]
colnames(TOC_data) <- c("station","depth", "Year_CE","TOC")
TOC_data <- TOC_data %>% filter(! is.na(station)) %>% filter(station != "EMB262_3_10_MUC")

#load biomarkers 

egb_gof_biomarker <- read.table( file=file.path(metadata_path,"Supplementary_Biomarker_data_GoF_EGB.tsv"), header=T, sep="\t")
egb_gof_biomarker <- egb_gof_biomarker[,grep("^C\\d|Dino*|age|location", colnames(egb_gof_biomarker))]

```


## create output paths
```{r}
workspace <- "04_workspace_2025"
output_path <- "04_results_2025"
output_path_extra <- file.path(output_path, "suppl")

if (!dir.exists(output_path)){ dir.create(output_path) }
if (!dir.exists(output_path_extra)){ dir.create(output_path_extra) }
if (!dir.exists(workspace)){ dir.create(workspace) }

```

#plot TOC data
```{r}
plot <- TOC_data%>% 
  mutate(location = ifelse(station == "EMB262_6_28_MUC", "Eastern Gotland Basin",
                    ifelse(station == "EMB262_13_8_MUC","Landsort Deep", "Gulf of Finland")))%>% 
  ggplot(aes(x= as.numeric(Year_CE), y= TOC))+ 
  geom_line()+facet_wrap(~location, scale="free_x")+ labs(x= "Years CE", y= "TOC in %")+ 
  figure_theme

print(plot)
ggsave(plot, file=file.path(output_path_extra, "MUC__TOC_values_over_time.png"),height=5, width=14)
ggsave(plot, file=file.path(output_path_extra, "MUC__TOC_values_over_time.tiff"),height=5, width=14, dpi=600)

TOC_data$Year_CE <- NULL
```


#define main figure layout
## approach colors
```{r}
## approach colors
hoiho2 <-c("#44781E","#1F59C7", "#9ABBD5", "#746455")
print_pal(hoiho2)
```

## eukaryotic colors
```{r}
#colors for eukaryotic
selected_colours <- c("#2B4B3C","#7D9D33", "#CED38C", "#DCC949", "#BCA888", "#CD8862", "#775B24") #

eukaryote <- c("#2B4B3C","#719134" ,"#B6C372" , "#BF964AFF", 
               "#C09E7D", "#F7C145FF","#1F6683", "#6A5557")

hoiho<- colorRampPalette(selected_colours)(8)
print_pal(hoiho)
print_pal(eukaryote)
#eukaryotic major communities
euka_taxonomic_groups <- c("Streptophyta","Phytoplankton","Macroalgae", "Fungi", 
                           "Rhizaria",  "Metazoa", "Opalozoa", "Other Protists",
                           "Unassigned") #n=8
euka_colors_taxo <- c(eukaryote, "darkgrey")
```

## define colors phytoplankton
```{r}

##phytoplankton -> not interested in Bacillariophyta
beyonce <- c( "#491C09FF", "#722D19FF", "#A54E21FF", "#CC842AFF", "#E7A92FFF", "#DFB54DFF", "#DDB855FF", "#F2E488FF")
greenish <- c( "#193E44FF", "#539E59FF", "#84C876FF", "#BBD961FF", "#F5EC54FF")
greenish<- colorRampPalette(greenish)(6)
taxonomic_groups <- c("Charophyta" ,"Chlorophyta" ,"Cryptophyta" ,"Haptophyta","Dinoflagellata" ,"Ochrophyta") 
colors_taxo <- greenish

```

# Prepare datasets
## Prepare biomarker data
```{r}
colnames(egb_gof_biomarker)
egb_gof_biomarker <- egb_gof_biomarker %>% 
        mutate(interval = cut(as.integer(age), breaks = seq(1835, 2025, by = 5), right = FALSE))

interval_egb_gof_biomarker <- egb_gof_biomarker %>% pivot_longer(!c(age, location, interval), names_to = "method", values_to = "values")%>%
  group_by(location, interval, method) %>% summarise(av_value = mean(values)) %>% na.omit() %>% 
  mutate(organism = ifelse(grepl("Dino", method),"Dinoflagellata", "Phytoplankton" ))

```

## Prepare Biomonitoring data

```{r}

filtered_ices_data <- ices_data %>%
        mutate( phylum = gsub("Myzozoa", "Dinoflagellata", phylum))%>%
        filter(phylum!="Cyanobacteria")%>% 
        filter(phylum != "Bacillariophyta")
```


```{r}
# average daily data
averaged_ices_data <- filtered_ices_data %>%
  group_by(MUC, Year, Month, AphiaID, phylum, order, family,genus, WoRMS_name, AphiaID_accepted)%>%
  summarise(averaged_value= mean(final_value), 
            median_value = median(final_value))

#average monthly data
summed_ices_data <- averaged_ices_data %>%
  group_by(MUC, Year,  phylum)%>%
  summarise(summed_value= sum(averaged_value)) 

#summarize data for plotting
plot_summed <- summed_ices_data %>% 
    pivot_wider(names_from = Year, values_from = summed_value, values_fill = 0)%>%
    pivot_longer(!c(MUC,  phylum), names_to = "Year", values_to = "summed_value") %>%
    group_by(MUC, Year)%>% ## remove 0 for years were just the record is missing
    filter(sum(summed_value) >0)

plot_supp_moni1 <- ggplot(plot_summed, aes(x= as.numeric(Year), y= summed_value, fill= phylum, group =phylum))+
    geom_area(color="grey25", size=0.1) + facet_wrap(~MUC, scales = "free") + figure_theme+
    labs(title="Biomonitoring data", x= "Years CE", y= "Count", fill="Phylum")+ 
    scale_fill_manual(breaks = taxonomic_groups,
                        values=colors_taxo)+
   scale_x_continuous(breaks = seq(1980, 2020, 5))

print(plot_supp_moni1)
#ggsave(plot_supp_moni1, file=file.path(output_path_extra, "biomonitoring_yearly_phytoplankton_over_time.png"), height=8, width=17)


moni_time <- plot_summed %>% group_by( MUC) %>% summarise(start=min(Year))
```

### Dinoflagellate data biomonitoring
```{r}
averaged_dino_data <- averaged_ices_data %>%
  filter(phylum =="Dinoflagellata")%>%
  group_by(MUC, Year,  phylum, order, family)%>%
  summarise(summed_value= sum(averaged_value))

plot_supp_moni2 <-averaged_dino_data%>%
  group_by(MUC, Year,  phylum, order)%>%
  summarise(summed_value= sum(summed_value)) %>%
  pivot_wider( names_from = "Year", values_from = "summed_value", values_fill=0)%>%
  pivot_longer(!c(MUC,phylum,order),names_to = "Year", values_to = "summed_value")%>%
  mutate(order= gsub("0", "NA", order))%>%
  ggplot( aes(x= as.numeric(Year), y= summed_value, fill= order, group =order))+
    geom_area() + facet_wrap(~MUC, scales = "free") + figure_theme+
    geom_vline(xintercept=2000, size= 0.7,linetype="dashed")+
    labs(title="Dinoflagellate community composition April to June", x= "Years CE", y= "Norm. counting data", fill= "Order")+ 
    scale_x_continuous(breaks = seq(min(averaged_dino_data$Year), max(averaged_dino_data$Year), 10))+
    scale_fill_viridis_d( option="magma")
print(plot_supp_moni2)
ggsave(plot_supp_moni2, file=file.path(output_path_extra, "Monitoring__dinoflagellate_community_composition__orderLevel_yearly.png"), height=8, width=16)
ggsave(plot_supp_moni2, file=file.path(output_path_extra, "Monitoring__dinoflagellate_community_composition__orderLevel_yearly.tiff"), height=8, width=16, dpi=600)
```


## Prepare Metabarcoding data

```{r}
#sum all reads for every replicate and average total reads of the replicates
total.reads <-seda_community %>%
        group_by(station, location, tag, replicate,depth, age)%>%
        summarise(total.reads.repl = sum(reads)) %>%
        #mean replicates
        group_by(station, location,  depth, age)%>%
        summarise(total.reads = mean(total.reads.repl)) %>%
        ungroup()


total.reads <-merge(total.reads, TOC_data, by = c("depth", "station"), all =T)%>% unique()

## calulcate the mean for every sample the average read number to average the replicates
seda_community.final <- seda_community %>%
        group_by(station, location, tag ,depth,  division, subdivision,class, order, family, genus, species  ) %>%
        summarise(mean.reads = mean(reads), 
                  replicates = sum(1*(reads >0)))%>%
        ungroup()

## average for the depth since several tags/samples for the same depth exist
seda_community.final <- seda_community.final %>%
        group_by(station, location,depth,  division, subdivision,class, order, family, genus, species  ) %>%
        summarise(mean.reads = mean(mean.reads), 
                  replicates = sum(replicates))%>%
        ungroup()
#rm(seda_community)

## categorize the taxonomic groups to make comparable with biomonitoring
seda_community.final_cat <- seda_community.final %>% # merge(seda_taxo, seda_community, by = "ID") %>% 
          replace(is.na(.),"unknown")%>% 
          mutate( taxon = ifelse((division =="Streptophyta"& class != "Zygnemophyceae"), "Streptophyta", 
                          ifelse(division== "Rhizaria", "Rhizaria",
                          ifelse(division== "Rhodophyta"| class =="Phaeophyceae"| class =="Ulvophyceae" , "Macroalgae",
                          ifelse(subdivision == "Fungi", "Fungi", 
                          ifelse(subdivision == "Metazoa", "Metazoa",
                          ifelse((subdivision == "Dinoflagellata"& class != "Syndiniales") | # 
                                  division =="Haptophyta" |
                                  (division == "Chlorophyta"& class !="Ulvophyceae")| division == "Prasinodermophyta" | # chlorophyta
                                  division =="Cryptophyta"| division =="Kathablepharidacea"| # cryptophyta
                                  class=="Bacillariophyceae"|
                                  class== "Coscinodiscophyceae" | class=="Mediophyceae"| class=="Chrysophyceae"|class=="Xanthophyceae"|  #ochrophyta
                                  (division =="Streptophyta"& class == "Zygnemophyceae")|
                                  (subdivision == "Gyrista" & class != "Pirsoniales" & class != "Peronosporomycetes" & !is.na(class)& 
                                               class != "Gyrista_X"		& class != 	"Hyphochytriomyceta" & class !="Phaeophyceae" )|
                                  subdivision =="Euglenozoa", "Phytoplankton",
                          ifelse(class=="Opalozoa","Opalozoa",
                          ifelse(division == "unknown", "Unassigned","Other Protists")))))))))%>%
        mutate(taxon_phyto = ifelse((subdivision == "Dinoflagellata" & class != "Syndiniales"), "Dinoflagellata",
                             ifelse(division=="Chlorophyta" | division == "Prasinodermophyta", "Chlorophyta", 
                             ifelse(division=="Haptophyta", "Haptophyta", 
                             ifelse(class == "Bacillariophyceae" ,  "Bacillariophyta",
                             ifelse(class == "MOCH-2" | class == "MOCH-3" | class == "MOCH-5" |class == "Eustigmatophyceae" | class == "Coscinodiscophyceae" | 
                                    class =="Mediophyceae"|  class =="Xanthophyceae"|class =="Bolidophyceae"| 
                                    class =="Chrysophyceae"| class =="Chrysomeridophyceae"|class =="Chrysomerophyceae"|
                                    class =="Dictyochophyceae"|class =="Pelagophyceae", "Ochrophyta",
                             ifelse((division =="Streptophyta"& class == "Zygnemophyceae"), "Charophyta",
                             ifelse(division=="Cryptophyta" |division =="Kathablepharidacea", "Cryptophyta", 
                             ifelse(subdivision=="Euglenozoa", "Euglenozoa", 
                             ifelse( is.na(class), division, class)))))))))) %>% unique()
write.table(seda_community.final_cat, 
            file = file.path(output_path, "Seda_community_data_categorized.tsv"), sep = "\t", row.names = F)

## overview 
seda_community.final_cat %>%
  group_by(taxon, taxon_phyto) %>% summarise(sum = n())

### add other protists?
total.reads.taxo <- seda_community.final_cat %>% # merge(seda_taxo, seda_community, by = "ID") %>% 
          group_by(station,  depth,  taxon, location)%>% #,division, subdivision, class
          summarise(total.reads.taxo = sum(mean.reads)) %>%
        ungroup()
          

#total.reads.taxo[is.na(total.reads.taxo)] <- "Unassigned"
total.reads.taxo <- total.reads.taxo %>%
  group_by(station, location, depth,  taxon) %>% 
  summarise(total.reads.taxo = mean(total.reads.taxo))%>%
        ungroup()  

total.reads.taxo <- merge(total.reads.taxo, total.reads, by = c("station", "depth", "location"))

total.reads.taxo$prop.taxon <- (total.reads.taxo$total.reads.taxo / total.reads.taxo$total.reads)*100

#rm(seda_community.final, TOC_data)      
```

###other protists? --> mainly Syndiniales
```{r}
seda_community.final_cat %>% filter(taxon=="Other Protists") %>% ungroup()%>% 
  select(division, subdivision, class)%>% unique() %>% arrange(division, subdivision, class)


other_protists <-seda_community.final_cat %>% filter(taxon=="Other Protists")%>% # merge(seda_taxo, seda_community, by = "ID") %>% 
          group_by(location, station,  depth, taxon, division, subdivision, class)%>% 
          summarise(total.reads.taxo = sum(mean.reads))%>%
          ungroup() %>%
          left_join(total.reads, by = c("station", "depth",  "location"))%>%
          mutate(prop.taxon = (total.reads.taxo / total.reads)*100)

# sort to get most prominent classes
other_protists %>% 
      arrange(desc(as.numeric(prop.taxon)))

```

### interval data of metabarcoding
```{r}
total.reads.taxo <- total.reads.taxo %>%
  mutate(interval = cut(as.numeric(age), breaks = seq(1790, 2025, by = 5), right = FALSE))  %>% 
  group_by(location, interval, taxon) %>%
  summarise(prop.taxon = mean(prop.taxon)) %>%
  mutate(year1 = gsub("\\[(\\d+),.*", "\\1",as.character(interval))) %>% unique()%>%
  mutate(year1=as.numeric(year1))

plot_meta_all <- ggplot(total.reads.taxo, aes(x=as.numeric(year1), y=prop.taxon, fill=taxon, group =taxon)) + 
    geom_area()+ 
    figure_theme+ 
    labs(title= "Metabarcoding - Community composition of eukaryotes", x= "5-year intervals CE", y= "Abundance proxy", fill= "Taxon") +
    facet_wrap(~location, nrow=1, scales = "free_x")+
    scale_x_continuous(breaks = seq(floor(min(total.reads.taxo$year1)/10)*10, ceiling(max(total.reads.taxo$year1)/10)*10, 20))+ 
    scale_fill_manual(breaks = euka_taxonomic_groups,
                        values=euka_colors_taxo)

print(plot_meta_all)
ggsave(plot_meta_all, file=file.path(output_path, "Metabarcoding_community_comp_over_time__interval.png"), height=8, width=22)
ggsave(plot_meta_all, file=file.path(output_path, "Metabarcoding_community_comp_over_time__interval.tiff"), height=8, width=22, dpi=600)
write.table(total.reads.taxo, file=file.path(output_path, "community_comp_over_time__interva.tsv"), row.names = F, sep ="\t")
```
## plot for supplementary


```{r}
class_community <-   seda_community.final_cat %>%
        group_by(station,location, depth, taxon, taxon_phyto, division, subdivision ,class) %>% 
        summarise(asv.richness = sum(1*(replicates >0)), 
                  total.read.mean = sum(mean.reads), 
                  biomass.equivalent = sum(replicates)/asv.richness) %>%
        group_by(station,location, depth, taxon, taxon_phyto, division, subdivision ,class) %>% 
        summarise(asv.richness = mean(asv.richness), 
                  total.read.mean = mean(total.read.mean), 
                  biomass.equivalent = mean(biomass.equivalent))


### extract Dinoflagellates
class_community.dino  <- class_community %>%
          filter((subdivision == "Dinoflagellata" & class != "Syndiniales"))


## extract phytoplankton
class_community.phyto  <- class_community %>%
          filter((subdivision == "Dinoflagellata" & class != "Syndiniales")|  # 
                 division =="Haptophyta" |
                 (division == "Chlorophyta"& class !="Ulvophyceae")|
                 division =="Cryptophyta"|division =="Kathablepharidacea"|
                 division == "Prasinodermophyta" |
                 (division =="Streptophyta"& class == "Zygnemophyceae")|
                 class == "Coscinodiscophyceae" | class =="Mediophyceae"| class=="Chrysophyceae"|class=="Xanthophyceae"|  #Gyrista 
                 (subdivision == "Gyrista" & class != "Pirsoniales" & class != "Peronosporomycetes" & class!= "unknown" & 
                       class != "Gyrista_X"		& class != 	"Hyphochytriomyceta" & class !="Phaeophyceae" )|
                 subdivision =="Euglenozoa" ) 



##### check impact on syndiniales (parasites)
class_community.phyto <- class_community.phyto %>% 
            filter(class != "Syndiniales")%>% ## filter because parasites
            filter(class!="Mediophyceae") %>%    ## filter because seaice specie/taxa
            filter(class !="Bacillariophyceae")   ##m filter because spring taxon
##########################


#rename taxo
class_community.phyto <- class_community.phyto %>% 
                            group_by(station, depth, division, taxon_phyto, location) %>% 
                            mutate(taxon =NULL)%>% 
                            summarise(asv.richness = sum(asv.richness), 
                                      total.read.mean = sum(total.read.mean))

class_community.phyto <- merge(class_community.phyto, total.reads, by=c("station", "depth",  "location"))
class_community.phyto$rel.abu <- (class_community.phyto$total.read.mean/class_community.phyto$total.reads) *100
```
```{r}
seda_community.final_cat[seda_community.final_cat =="unknown"] <- NA
phytoplankton_identified <- seda_community.final_cat%>%
        filter(taxon =="Phytoplankton")%>%
        filter(class!= "Bacillariophyceae" & class!="Mediophyceae")%>%
        mutate(identified= ifelse(!is.na(species), species,
                            ifelse(!is.na(genus), genus,
                            ifelse(!is.na(family), family, 
                            ifelse(!is.na(class), class, subdivision)))))%>%
        group_by(station,location, depth, division, subdivision ,class, family, genus,species, identified) %>% 
        summarise(asv.richness = sum(1*(replicates >0)), 
                  total.read.mean = sum(mean.reads), 
                  biomass.equivalent = sum(replicates)/asv.richness) %>%
        left_join(total.reads, by=c("station", "depth",  "location"))%>%
        mutate(rel.abu = (total.read.mean/total.reads) *100)%>%
        mutate(interval = cut(as.numeric(age), breaks = seq(1790, 2025, by = 5), right = FALSE))  %>%
        mutate(year1 = gsub("\\[(\\d+),.*", "\\1",as.character(interval))) %>%
        mutate(year1= as.numeric(year1))%>%ungroup()


## define taxa of interest
 phytoplankton_identified.plot <- phytoplankton_identified %>% 
  filter(division== "Stramenopiles"|division== "Alveolata"|division== "Chlorophyta") %>%
     mutate(class= ifelse(!is.na(class), class, division))%>%
    mutate(year1= as.numeric(year1))

#  set colors and alpha values
colors_common <- c("#EE9A49", "darkseagreen4", "#00688B")

color_shades_division <- phytoplankton_identified.plot %>% 
  group_by(division, class) %>%
  select(division, class) %>% unique() %>%ungroup()

# arrange is not properly workding
color_shades_division <- color_shades_division[order(color_shades_division$division, color_shades_division$class),] %>%
  rename(alpha_name= class, color_name= division) %>%
  mutate(col_fa= factor(color_name,levels = unique(color_name)),
         col_no = as.numeric(col_fa))%>%
  group_by(col_no)%>%
  mutate(alpha =seq( (1/n()), 1,length.out = n()))%>%
  select(alpha_name, col_no, alpha)

phytoplankton_identified.plot <- phytoplankton_identified.plot %>% 
  group_by(location, interval, year1, division, class) %>%
  summarise(values = sum(rel.abu)) %>% unique()%>%ungroup()

plot_suppl_metaAbu1 <-  ggplot(phytoplankton_identified.plot, 
                               aes(x=year1, y=values, fill=division, group = class, alpha=factor(class, levels= color_shades_division$alpha_name))) +
                          geom_area( size=0.2, colour="black")+ 
                          geom_vline(xintercept=2000, size= 0.7,linetype="dashed")+
                          labs(title= "Planktonic classes", x= "5-year intervals CE", y= "Abundance proxy", alpha= "Class")+
                          figure_theme+ 
                          scale_x_continuous(breaks = seq(min(phytoplankton_identified.plot$year1), max(phytoplankton_identified.plot$year1), 10))+ 
                          facet_grid(division~location, scales = "free_y")+
                        # for shades not hues
                          scale_fill_manual(values = colors_common) +
                          scale_alpha_manual(values = color_shades_division$alpha)+
                          ## add colors to legend
                          guides(fill = guide_none(),
                                alpha = guide_legend(override.aes = list(fill = colors_common[color_shades_division$col_no]))) 
print(plot_suppl_metaAbu1)
ggsave(plot_suppl_metaAbu1, file=file.path(output_path_extra, "Metabarocding__community_comp_over_time__interval_abuClass.png"), height=8, width=16)
ggsave(plot_suppl_metaAbu1, file=file.path(output_path_extra, "Metabarocding__community_comp_over_time__interval_abuClass.tiff"), height=8, width=16, dpi=600)

```


```{r}

intervall_metabarcoding_relabu <- class_community.phyto %>%
  mutate(interval = cut(as.numeric(age), breaks = seq(1790, 2025, by = 5), right = FALSE))  %>% 
  group_by(location, interval, taxon_phyto) %>%
  summarise(values = mean(rel.abu), richness= mean(asv.richness)) %>%
  mutate(year1 = gsub("\\[(\\d+),.*", "\\1",as.character(interval))) %>% unique()%>% 
  mutate(year1 =as.numeric(year1))

unique(intervall_metabarcoding_relabu$taxon_phyto)

##plotting full muc relative abundance
moni_time <- moni_time %>% mutate(start = as.numeric(start))  %>% dplyr::rename(location = MUC)
  
plot_supp_meta2 <- intervall_metabarcoding_relabu  %>%
  ggplot( aes(x=year1, y=values, fill=taxon_phyto, group = taxon_phyto)) +
    geom_area(color="grey25", size=0.1)+ 
    geom_vline(data=moni_time, aes(xintercept=start), size= 0.7,linetype="dashed")+
    labs(title= "Metabarcoding - Community composition of phytoplankton ", x= "5-year intervals CE", y= "Abundance proxy", fill ="Taxon")+
    figure_theme+ 
    scale_fill_manual(breaks = taxonomic_groups, values=colors_taxo)+
    scale_x_continuous(breaks = seq(floor(min(intervall_metabarcoding_relabu$year1)/10)*10, ceiling(max(intervall_metabarcoding_relabu$year1)/10)*10, 20))+ 
    facet_grid(~location, scales= "free_x")+
    theme(legend.position = "None")
print(plot_supp_meta2)

plot <- ggarrange(plot_meta_all, plot_supp_meta2 , plot_supp_moni1, 
                  nrow = 3, labels = c("A","B", "C"),
                  heights = c(1,0.8,1),
                  font.label = list(size = 26, color = "black", face = "bold", family = NULL))


print(plot)
ggsave(plot,  file=file.path(output_path, "community_comp_phytoplankton_ALL.png"), height=14, width=14)
ggsave(plot,  file=file.path(output_path, "community_comp_phytoplankton_ALL.tiff"), height=14, width=14, dpi=600)

```

# MODELING

## prepare & combine data

```{r}
total.reads.taxo.original <- total.reads.taxo


### prepare data for k determiantion
total.reads.yearly <- seda_community.final_cat %>% # merge(seda_taxo, seda_community, by = "ID") %>% 
          rename(organism=taxon)%>%
          filter(organism == "Phytoplankton") %>%
          filter(class!= "Bacillariophyceae" & class!="Mediophyceae")

### dinoflagellates in detail
total.reads.yearly.dino.community <- total.reads.yearly %>% 
          filter(subdivision == "Dinoflagellata") %>% mutate(organism ="Dinoflagellata") %>%
          #mutate(family= ifelse(!is.na(family), family,ifelse(! is.na(order), order, class))) %>%
          group_by(location,  depth,  organism, class, order, family)%>% #,division, subdivision, class
          summarise(total.reads.taxo = sum(mean.reads)) 

total.reads.yearly.dino.community <-merge(total.reads, total.reads.yearly.dino.community, by = c( "location", "depth")) %>% 
              #rename("values" = total.reads.taxo) %>% 
              mutate(method=rep("Metabarcoding", length(age)), age = as.character(age),
                     values=as.numeric(total.reads.taxo)/as.numeric(total.reads) *100)%>% 
              mutate(depth  = NULL, station =NULL, 
                     total.reads.taxo=NULL,total.reads=NULL, TOC= NULL) 

### dinoflagellates & phytonplanton in general
total.reads.yearly.dino <- total.reads.yearly %>% 
          filter(subdivision == "Dinoflagellata") %>% mutate(organism ="Dinoflagellata") %>%
          group_by(location,  depth, organism)%>% #,division, subdivision, class
          summarise(total.reads.taxo = sum(mean.reads)) 

total.reads.yearly.phyto <- total.reads.yearly %>%
          group_by(location, depth, organism)%>% #,division, subdivision, class
          summarise(total.reads.taxo = sum(mean.reads)) 

total.reads.yearly <- rbind(total.reads.yearly.dino, total.reads.yearly.phyto)
rm(total.reads.yearly.dino, total.reads.yearly.phyto)

total.reads.yearly <-merge(total.reads, total.reads.yearly, by = c("location", "depth")) %>% 
              mutate(method=rep("Metabarcoding", length(age)), age = as.character(age),
                     values=as.numeric(total.reads.taxo)/as.numeric(total.reads) *100)%>% 
              mutate( depth  = NULL, station =NULL, 
                     total.reads.taxo=NULL,total.reads=NULL, TOC= NULL) 
              
#### Biomonitoring data 
ices_data_yearly <- summed_ices_data %>% pivot_wider( names_from = phylum, values_from = summed_value, values_fill = 0 ) %>% 
            mutate( Phytoplankton = Chlorophyta+Cryptophyta+Dinoflagellata+Haptophyta+Ochrophyta+Charophyta) %>% 
            pivot_longer(!c(MUC, Year), names_to = "organism", values_to = "total.values") %>% 
            filter(organism == "Phytoplankton"|organism =="Dinoflagellata") %>% # organism=="Phytoplankton_Cyano"
            rename( "location" = MUC, "age" = Year)%>% mutate(method= rep("Biomonitoring", length(interval))) #"Biomonitoring, count data"


max_counts <- ices_data_yearly %>%
  group_by(organism, location) %>%
  summarize(overall_max_count = max(total.values, na.rm = TRUE))

ices_data_yearly_norm <- left_join(ices_data_yearly, max_counts, by = c("organism", "location"))

ices_data_yearly_norm <- ices_data_yearly_norm %>% mutate(values= (total.values/overall_max_count)*100)%>%
                                         mutate(total.values = NULL, overall_max_count= NULL)

##### Biomarker data
egb_gof_biomarker.yearly <- egb_gof_biomarker %>% mutate(interval=NULL) %>% pivot_longer(!c(age, location), names_to = "method", values_to = "real.conc")%>%
  mutate(organism = ifelse(grepl("Dino", method),"Dinoflagellata", "Phytoplankton" ))%>% 
  mutate(location = as.factor(location), method = as.factor(method), organism = as.factor(organism), age = as.character(age))


max_counts <- egb_gof_biomarker.yearly %>%
  group_by(method, location) %>%
  summarize(overall_max_conc = max(real.conc, na.rm = TRUE))

egb_gof_biomarker.yearly_norm <- left_join(egb_gof_biomarker.yearly, max_counts, by = c("method", "location"))

egb_gof_biomarker.yearly_norm <- egb_gof_biomarker.yearly_norm %>% mutate(values= (real.conc/overall_max_conc)*100)%>%
                                         mutate(real.conc = NULL, overall_max_conc= NULL)

### combine all three datasets normalized
total_data.yearly_norm <- rbind( ices_data_yearly_norm, egb_gof_biomarker.yearly_norm)
total_data.yearly_norm <- rbind( total_data.yearly_norm, total.reads.yearly)
total_data.yearly_norm <- total_data.yearly_norm %>% mutate( location2 = ifelse(location == "Eastern Gotland Basin", "EGB", 
                                                               ifelse(location == "Landsort Deep", "LD",
                                                               ifelse(location == "Gulf of Finland", "GoF", NA))))

### orginal data 
egb_gof_biomarker.yearly <- egb_gof_biomarker.yearly %>%
      dplyr::rename(values=real.conc)

ices_data_yearly <- ices_data_yearly %>%
  dplyr::rename(values=total.values )

total_data.yearly_real<- rbind( ices_data_yearly, egb_gof_biomarker.yearly)
total_data.yearly_real <- rbind( total_data.yearly_real, total.reads.yearly)
total_data.yearly_real <- total_data.yearly_real %>% mutate( location2 = ifelse(location == "Eastern Gotland Basin", "EGB", 
                                                               ifelse(location == "Landsort Deep", "LD",
                                                               ifelse(location == "Gulf of Finland", "GoF", NA))))
```
note: Gymnodiniaceae most abund ant after NA
```{r}
total.reads.yearly.dino.community <- total.reads.yearly.dino.community %>%mutate(age= as.numeric(age))  
colnames(total.reads.yearly.dino.community)
plot_supp_meta3 <- total.reads.yearly.dino.community %>% 
  group_by(location, age, organism, class, order)%>%
  summarise(values=sum(values))%>%
  ggplot( aes(x=age, y=values,  group =order, fill =order))+
  geom_area(aes())+ figure_theme + facet_wrap( ~ location) + 
  geom_vline(xintercept=2000, size= 0.7,linetype="dashed")+
  scale_x_continuous(breaks = seq(floor(min(total.reads.yearly.dino.community$age)/10)*10, ceiling(max(total.reads.yearly.dino.community$age)/10)*10, 20))+ 
  labs( title= "Relative abundance of dinoflagellate orders over time", x= "Years CE", y= "Abundance proxy", fill ="Order")
print(plot_supp_meta3)
#ggsave(plot_supp_meta3, file=file.path(output_path_extra, "Metabarcoding_dino_communities_classes__MUC.png"), height=7, width=17)

plot_supp_meta4 <- total.reads.yearly.dino.community %>% 
  select(- order)%>%
  group_by(location, age, organism, class, family)%>%
  summarise(values=sum(values))%>%
  ggplot( aes(x=age, y=values,  group =family, fill =family))+
  geom_area()+ figure_theme + facet_wrap( ~ location) + 
  geom_vline(xintercept=2000, size= 0.7,linetype="dashed")+
  scale_x_continuous(breaks = seq(floor(min(total.reads.yearly.dino.community$age)/10)*10, ceiling(max(total.reads.yearly.dino.community$age)/10)*10, 20))+ 
  labs( title= "Relative abundance of dinoflagellate families over time", x= "Years CE", y= "Abundance proxy", fill ="Family")

print(plot_supp_meta4)
#ggsave(plot_supp_meta4, file=file.path(output_path_extra, "Metabarcoding_dino_communities_GymnodiniaceaeafterNA__MUC.png"), height=7, width=17)
rm(total.reads.yearly.dino.community)
```
####
# save data for next steps 
```{r}
save(total_data.yearly_norm, file=file.path(workspace,"Data_allApproach__for_modelling.RData"))
save(total_data.yearly_real, file=file.path(workspace,"Data_allApproach__for_modelling__real.RData"))
save.image(file=file.path(workspace,"WorkSpace_Mainscript.RData"))

```


##Create Suppl
```{r}

plot_suppl_meta <- ggarrange(  plot_supp_meta3, plot_supp_meta4, nrow = 2, labels = c("A","B"),
           font.label = list(size = 26, color = "black", face = "bold", family = NULL))
print(plot_suppl_meta)


ggsave(plot_suppl_meta,  file=file.path(output_path, "Supplementary__metabarcoding_results.png"), height=16, width=18)
ggsave(plot_suppl_meta,  file=file.path(output_path, "Supplementary__metabarcoding_results.tiff"), height=16, width=18, dpi=600)

```


